Doing Better Than UCT: \\ Rational Monte Carlo Sampling in Trees

### 2011-08-17 paper/rmcts/paper.tex
some fixes
some fixes after eyals notes

\begin{abstract}
UCT, a state-of-the art algorithm for Monte Carlo tree sampling
(MCTS), is based on UCB, a sampling policy for the Multi-armed Bandit
Problem (MAB) that minimizes the accumulated regret. However, MCTS
differs from MAB in that only the final choice, rather than all arm
pulls, brings a reward, that is, the simple regret, as opposite to the
accumulated regret, must be minimized. This work introduces policies for
multi-armed bandits with lower simple regret than UCB, and an
algorithm for MCTS which combines accumulated and simple regret
minization and outperforms UCT. Finite-time and asymptotic analysis of
the policies is provided, and the algorithms are empirically compared.
\end{abstract}

@@

Such strategies are used to select the best arm, and, by extension,
the best action in Monte Carlo tree search. Monte Carlo tree search is
used to solve Markov Decision Processes (MDP) approximately. An MDP is
defined by the set of states $S$, the set of actions $A$, the
transition table $T(s, a, s')$, the reward table $R(s, a, s')$, the
initial state $s$ and the goal state $t$: $(S, A, T, R, s, t)$
\cite{Russell.aima}.  Monte Carlo tree search explores an MDP by
\emph{rollouts}---trajectories from the current state to a state in
which a termination condition is satisfied (either the goal state, or
a cutoff state for which the reward is evaluated
approximately). Algorithm UCB that maximizes the accumulated reward
\cite{Auer.ucb} had been extended to tree search as UCT
\cite{Kocsis.uct}.

UCT-driven search achieves good results in many search problems. At each
search step, the algorithm allocates Monte Carlo rollouts to choose
the best action, thereby minimizing the simple regret. However, the
simple and the accumulated regret cannot be minimized simultaneously;
moreover, \cite{Bubeck.pure} shows that in many cases the smaller the
accumulated regret, the greater is the simple regret.

UCT performance can be improved by combining UCB with a sampling
scheme that minimizes the simple regret of selecting an action at the
current root node. Indeed, the algorithm must select an action with
the minimum regret \emph{given assumption that after performing the selected action
the algorithm performs optimally}, which corresponds to maximizing the
value of partial information \cite{Russell.aima}. Therefore, an
improved allocation scheme would
\begin{itemize}
\item maximize the value of partial
information by sampling actions to minimize the \textbf{simple} regret at the
current root node, and then 
\item minimize the \textbf{accumulated} regret of the rollouts from the second
  step on to approach the optimum behavior. 
\end{itemize}

@@

Such sampling scheme is significantly less sensitive to the tuning of
the exploration factor ($C_p$ in \cite{Kocsis.uct}) of UCT, since the
conflict \cite{Bubeck.pure} between the need for a larger value of
$C_p$ on the first step (simple regret) and a smaller value for the
rest of the rollout (accumulated regret) is resolved. In fact, a
sampling scheme that uses UCB at all steps but a larger value of $C_p$
for the first step than for the rest of the steps, outperforms
UCT. The pseudocode of the two-stage rollout is in
Algorithm~\ref{alg:two-stage-mcts}.

\begin{algorithm}[t]
\caption{Two-stage Monte-Carlo tree search sampling}
\label{alg:two-stage-mcts}
\begin{algorithmic}[1]
\Procedure{Rollout}{node, depth=1}
  \If {\Call{IsLeaf}{node, depth}}
    \State \textbf{return} 0
  \Else
    \If {depth=1}
      \State action $\gets$ \Call{FirstAction}{node} \Comment minimizes simple regret, e.g. $\frac 1 2$-greedy
    \Else
      \State action $\gets$ \Call{NextAction}{node} \Comment  UCB --- minimizes accumulated regret
    \EndIf
    \State next-node $\gets$ \Call{NextState}{node, action}
    \State reward $\gets$ \Call{TransitionReward}{node, action, next-node}
    \State \hspace{4em} + \Call{Rollout}{next-node, depth+1} \Comment call {\sc Rollout} recursively
    \State \Call{UpdateStats}{node, action, reward}
  \EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

Figure~\ref{fig:mab-simple-regret}.b shows the regret
vs. the number of samples, averages over $10^4$ experiments, for uniform
random sampling (RND), UCT (same as UCB in this case),
$\frac 1 2$-greedy (GCT), and QCT (UCT with $\log(\cdot)$ replaced by
$\sqrt{\cdot}$), for randomly generated instances of 32 arms.

@@

Figure~\ref{fig:mcts-regret} shows a sketch of the search tree
(Figure~\ref{fig:mcts-regret}.a) and the dependency of the regret vs. the
number of samples for trees with root degree 16
(Figure~\ref{fig:mcts-regret}.b) and 64 (Figure~\ref{fig:mcts-regret}.c). The
dependencies look differently from Multi-armed bandit instances, but
the algorithms exhibit a similar relative performance: GCT and QCT
simultaneously are better than UCT, QCT dominates UCT everywhere
except for small numbers of instances. The advantage of QCT grows with
the number of arms.

@@

Figure~\ref{fig:sailing-cost-vs-factor} shows the rerget vs. the
factor for different numbers of samples. QCT is always better than
UCT, and $\frac 1 2$-greedy is better than UCT expect for a small range of
values of the exploration factor $C_p$.

@@

The Monte Carlo tree search algorithms presented in the paper differ
from UCT at the first step of the rollout, when the `simple' selection
regret is minimized instead of the accumulated regret. Both the
theoretical analysis and the empirical evaluation provide evidence for
better general performance of the proposed algorithms.

@@

@@

Bounds (\ref{eq:deriv-epsilon-greedy-b}, \ref{eq:deriv-ucb-alpha-b}) were
derived ignoring dependency of the allocation on expected values
of non-optimal arms. Derivation of better bounds seems
complicated. Instead, one can analyse symptotic behavior of each of the
algorithms in assumption that the sample mean is close to the expectation.

@@

Replacing the logarithm with a faster growing subpolynomial
function, like square root (UQB), results in a superpolynomial convergence
rate of the upper bound:
\begin{equation}
P_j^{uqb}\le\mathbb{B}P_j^{uqb}\approx2\exp\left(-\beta\Delta_j^2\frac {\sqrt n}{\Delta_j^2}\right)
            =2\exp(-\beta\sqrt n)
 \end{equation}
where $\beta$ is a constant that depends on reward bounds only.
Finite time convergence can still be shown similarly to UCB.

>>>>

\begin{abstract}
UCT, a state-of-the art algorithm for Monte Carlo tree sampling
(MCTS), is based on UCB, a sampling policy for the Multi-armed Bandit
Problem (MAB) that minimizes the accumulated regret. However, MCTS
differs from MAB in that only the final choice, rather than all arm
pulls, brings a reward, that is, the simple regret, as opposite to the
cumulative regret, must be minimized. This work introduces policies for
multi-armed bandits with lower simple regret than UCB, and an
algorithm for MCTS which combines cumulative and simple regret
minization and outperforms UCT. Finite-time and asymptotic analysis of
the policies is provided, and the algorithms are empirically compared.
\end{abstract}

@@

Such strategies are used to select the best arm, and, by extension,
the best action in Monte Carlo tree search. Monte Carlo tree search is
used to solve Markov Decision Processes (MDP) approximately. An MDP is
defined by the set of states $S$, the set of actions $A$, the
transition table $T(s, a, s')$, the reward table $R(s, a, s')$, the
initial state $s$ and the goal state $t$: $(S, A, T, R, s, t)$
\cite{Russell.aima}.  Monte Carlo tree search explores an MDP by
\emph{rollouts}---trajectories from the current state to a state in
which a termination condition is satisfied (either the goal state, or
a cutoff state for which the reward is evaluated
approximately). Algorithm UCB that minimizes the cumulative regret
\cite{Auer.ucb} had been extended to tree search as UCT
\cite{Kocsis.uct}.

UCT-driven search achieves good results in many search problems. At each
search step, the algorithm allocates Monte Carlo rollouts to choose
the best action, thereby minimizing the cumulative regret. However, the
simple and the cumulative regret cannot be minimized simultaneously;
moreover, \cite{Bubeck.pure} shows that in many cases the smaller the
cumulative regret, the greater the simple regret.

UCT performance can be improved by combining UCB with a sampling
scheme that minimizes the simple regret of selecting an action at the
current root node. Indeed, the algorithm must select an action with
the minimum regret \emph{given the assumption that after performing the selected action
the algorithm performs optimally}, which corresponds to maximizing the
value of partial information \cite{Russell.aima}. Therefore, an
improved allocation scheme would
\begin{itemize}
\item maximize the value of partial
information by sampling actions to minimize the \textbf{simple} regret at the
current root node, and then 
\item minimize the \textbf{cumulative} regret of the rollouts from the second
  step on to approach the optimum behavior. 
\end{itemize}

@@

Such sampling scheme is significantly less sensitive to the tuning of
the exploration factor ($C_p$ in \cite{Kocsis.uct}) of UCT, since the
conflict \cite{Bubeck.pure} between the need for a larger value of
$C_p$ on the first step (simple regret) and a smaller value for the
rest of the rollout (cumulative regret) is resolved. In fact, a
sampling scheme that uses UCB at all steps but a larger value of $C_p$
for the first step than for the rest of the steps, outperforms
UCT. The pseudocode of the two-stage rollout is in
Algorithm~\ref{alg:two-stage-mcts}.

\begin{algorithm}[t]
\caption{Two-stage Monte-Carlo tree search sampling}
\label{alg:two-stage-mcts}
\begin{algorithmic}[1]
\Procedure{Rollout}{node, depth=1}
  \If {\Call{IsLeaf}{node, depth}}
    \State \textbf{return} 0
  \Else
    \If {depth=1}
      \State action $\gets$ \Call{FirstAction}{node} \Comment minimizes simple regret, e.g. $\frac 1 2$-greedy
    \Else
      \State action $\gets$ \Call{NextAction}{node} \Comment  UCB --- minimizes cumulative regret
    \EndIf
    \State next-node $\gets$ \Call{NextState}{node, action}
    \State reward $\gets$ \Call{TransitionReward}{node, action, next-node}
    \State \hspace{4em} + \Call{Rollout}{next-node, depth+1} \Comment call {\sc Rollout} recursively
    \State \Call{UpdateStats}{node, action, reward}
  \EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

Figure~\ref{fig:mab-simple-regret}.b shows the regret
vs. the number of samples, averaged over $10^4$ experiments, for uniform
random sampling (RND), UCT (same as UCB in this case),
$\frac 1 2$-greedy (GCT), and QCT (UCT with $\log(\cdot)$ replaced by
$\sqrt{\cdot}$), for randomly generated instances of 32 arms.

@@

Figure~\ref{fig:mcts-regret} shows a sketch of the search tree
(Figure~\ref{fig:mcts-regret}.a) and the dependency of the regret vs. the
number of samples for trees with root degree 16
(Figure~\ref{fig:mcts-regret}.b) and 64 (Figure~\ref{fig:mcts-regret}.c). The
dependencies look differently from Multi-armed bandit instances, but
the algorithms exhibit a similar relative performance: either GCT or QCT
is always better than UCT, QCT dominates UCT everywhere
except for small numbers of instances. The advantage of QCT grows with
the number of arms.

@@

Figure~\ref{fig:sailing-cost-vs-factor} shows the rerget vs. the
exploration factor for different numbers of samples. QCT is always better than
UCT, and $\frac 1 2$-greedy is better than UCT expect for a small range of
values of the exploration factor $C_p$.

@@

The Monte Carlo tree search algorithms presented in the paper differ
from UCT at the first step of the rollout, when the `simple' selection
regret is minimized instead of the cumulative regret. Both the
theoretical analysis and the empirical evaluation provide evidence for
better general performance of the proposed algorithms.

@@

%TODO: bounds on regret based on information gathered, independent of
%the way the choice is made.

@@

%TODO: Oracle instead of asymptotic

Bounds (\ref{eq:deriv-epsilon-greedy-b}, \ref{eq:deriv-ucb-alpha-b}) were
derived ignoring dependency of the allocation on expected values
of non-optimal arms. Derivation of better bounds seems
complicated. Instead, one can analyse the asymptotic behavior of each of the
algorithms under the assumption that the sample mean is close to the expectation.

@@

Replacing the logarithm with a faster growing subpolynomial
function, like square root (UQB), results in a superpolynomial convergence
rate of the upper bound:
\begin{equation}
P_j^{uqb}\le\mathbb{B}P_j^{uqb}\approx2\exp\left(-\beta\Delta_j^2\frac {\sqrt n}{\Delta_j^2}\right)
            =2\exp(-\beta\sqrt n)
 \end{equation}
where $\beta$ is a constant that depends only on reward bounds.
Finite time convergence can still be shown similarly to UCB.
